---
title: "R Notebook"
output: html_notebook
---

This notebook is used to extract data from original data files in the Excel format. 

New run with the 2025 data: 2025-11-13
Time columns should be defined as "text" in list.col.types.inshore.

Edited for 2024. 2024-08-13
Need to update a few constants in GrayWhaleCalfProduction_fcns.R: list.sheet.names.inshore,
list.col.types.inshore, list.col.names.inshore.


```{r setup}
rm(list=ls())
library(tidyverse)
library(lubridate)
library(readxl)
source("GrayWhaleCalfProduction_fcns.R")

data.dir <- "data/All data/"
#out.dir <- "data/Processed data/"    # completed in May 2022 but needs revisions

# v3 includes the new shift definitions
ver <- "v3"  # v2

out.dir <- paste0("data/Processed data ", ver, "/")  # for revising outputs
if (!dir.exists(out.dir)) dir.create(out.dir)

out.formatted.dir <- paste0("data/Formatted Annual Data ", ver, "/")
if (!dir.exists(out.formatted.dir)) dir.create(out.formatted.dir)

# Regardless of if output file exists, this will overwrite the output files
save.files <- TRUE
start.time <- "100"   # This is for v3. For v2, should be year specific.


```

Event code:
1 = START EFFORT
2 = CHANGE OBSERVERS
3 = CHANGE SIGHTING CONDITIONS
4 = GRAY WHALE SIGHTING
5 = END EFFORT
6 = OTHER SPECIES SIGHTING

2023-06-01 A HA MOMENT. RATHER THAN MAKING ARBITRARY STARTING TIME OF 0630 OR 0700, CREATE 3-HR SHIFTS STARTING FROM 0100. 

NEW FUNCTIONS WERE CREATED AND TESTED. THE NEXT CHUNK WILL RUN ALL YEARS. 

```{r}
years <- c(1994:2019, 2021:2025)

# file index indicates which one of "inshore" and "offshore" file should be selected. In some years
# there were multiple files for each. No offshore files exist since 2017 so NA is in place
inshore.xls.file.idx <- c(1,1,1,1,1,1,1,1,1,1,2,1,1,1,1,1,2,2,2,1,1,1,1,1,1,1,1,1,1,1,1)
offshore.xls.file.idx <- c(2,2,2,2,2,2,2,2,2,2,3,2,2,2,2,2,3,3,3,2,2,3,2,NA,NA,NA,NA,NA,NA,NA,NA)

if (length(inshore.xls.file.idx) != length(years)) stop("The length of inshore file index does not match with that of years")

if (length(offshore.xls.file.idx) != length(years)) stop("The length of offshore file index does not match with that of years")


for (k in 1:length(years)){
  Year <- years[k]
  dir.name <- paste0("C_C ", Year)
  
  out.file.name.inshore <- paste0(out.dir, "Processed_inshore_data_",
                              Year, "_", ver, ".csv")
  
  if (!file.exists(out.file.name.inshore)){
    xls.files <- list.files(path = paste0(data.dir, dir.name), pattern = ".xls")
    xls.file.name.inshore <- paste0(data.dir, dir.name, "/", xls.files[inshore.xls.file.idx[k]])
    if (!is.na(offshore.xls.file.idx[k])){
      xls.file.name.offshore <- paste0(data.dir, dir.name, "/", xls.files[offshore.xls.file.idx[k]])
      
    }
    
    idx.list <- which(names(list.sheet.names.inshore) == paste0("Y", Year))
    if (Year < 2017){
      out.list <- get.all.data(list.sheet.names.inshore[[idx.list]], 
                               list.sheet.names.offshore[[idx.list]],
                               list.col.types.inshore[[idx.list]], 
                               list.col.types.offshore[[idx.list]],
                               list.col.names.inshore[[idx.list]], 
                               list.col.names.offshore[[idx.list]],
                               Year, 
                               xls.file.name.inshore, xls.file.name.offshore,
                               start.time)
      
    } else {
      out.list <- get.data.inshore.only(Year = Year,
                                        xls.file.name = xls.file.name.inshore,
                                        sheet.name = list.sheet.names.inshore[[idx.list]],
                                        col.types = list.col.types.inshore[[idx.list]],
                                        col.names = list.col.names.inshore[[idx.list]],
                                        start.time = start.time)
    }
    
    
    if (save.files){
      # Creates file names and saves files
      files <- file.names(out.dir = out.dir, 
                          ver = ver, 
                          Year = Year, 
                          out.list = out.list)
      
    }
    
    
  }
  
  
}

```

